---
title: "GWAS Analysis & Interactive plots"
author: "Bruno Ambrozio"
date: "11/10/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```
## Description 
* A [GWAS](https://en.wikipedia.org/wiki/Genome-wide_association_study) analysis, with post-analytic visualization and interrogation.
    * Analysis steps derived from the paper: ["A guide to genome-wide association analysis and post-analytic interrogation" (doi: 10.1002/sim.6605)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5019244/) (Reed et al., 2015).
    * Dataset from the PennCATH study of coronary artery disease (CAD). Paper: ["Identification of ADAMTS7 as a novel locus for coronary atherosclerosis and association of ABO with myocardial infarction in the presence of coronary atherosclerosis: two genome-wide association studies." (doi: 10.1016/S0140-6736(10)61996-4)](http://www.ncbi.nlm.nih.gov/pubmed/21239051) ((Reilly et al., 2011))
    * Detailed tutorial can be found [here](http://www.stat-gen.org/tut/tut_intro.html)

## Summary

1. Download necessary R packages and setting global parameters to save progress while working through the GWA analysis.
1. Include quality control steps for both SNP and sample level filtering. * [PCA](https://en.wikipedia.org/wiki/Principal_component_analysis) for [population stratification](https://en.wikipedia.org/wiki/Population_stratification) in statistical modeling, as well as imputation of non-typed SNPs using 1000 Genomes reference genotype data.
1. GWAS analysis' strategies.
    * Basic linear modeling functionality.
    * Imputed data using functionality contained.
1. Post-analytic interrogation.
    * Performance of statistical models.
    * Visualization of the global and subsetted GWAS output.
```{r}
# Configuring global parameters
# ===========================================================

# Set workdir
setwd("~/workspace/github.com/bambrozio/bioinformatics/gwas")

# location of the files to be downloaded/saved throughout the GWAS
data.dir <- '/Users/bambrozi/Downloads/gwas_cad_data'
out.dir <- data.dir # may want to write to a separate dir to avoid clutter

# Download files
urlSupport <- "https://www.mtholyoke.edu/courses/afoulkes/Data/GWAStutorial/GWASTutorial_Files.zip"
zipSupport.fn <- sprintf("%s/GWAStutorial_Files.zip", data.dir) 

# Input files
gwas.fn <- lapply(
  c(bed='bed', bim='bim', fam='fam', gds='gds'),
  function(n)
    sprintf("%s/GWAStutorial.%s", data.dir, n))

clinical.fn <- sprintf("%s/GWAStutorial_clinical.csv", data.dir) 

onethou.fn <- lapply(c(info='info',ped='ped'),
                     function(n)
                       sprintf("%s/chr16_1000g_CEU.%s", data.dir, n))

protein.coding.coords.fname <- sprintf("%s/ProCodgene_coords.csv",
                                       data.dir)

# Output files
gwaa.fname <- sprintf("%s/GWAStutorialout.txt", out.dir)
gwaa.unadj.fname <- sprintf("%s/GWAStutorialoutUnadj.txt", out.dir)
impute.out.fname <- sprintf("%s/GWAStutorial_imputationOut.csv", out.dir)
CETP.fname <- sprintf("%s/CETP_GWASout.csv", out.dir)

# Working data saved between each code snippet so each can run independently.
# Use save(data, file=working.data.fname(num))
working.data.fname <- function(num)
  { sprintf("%s/working.%s.Rdata", out.dir, num) }
```

## Preliminary
* Install packages:
    * [Bioconductor](http://www.bioconductor.org/).
        * [snpStats](http://www.bioconductor.org/packages/release/bioc/html/snpStats.html).
            * Read in various formats of genotype data.
            * Qality control.
            * Imputation and association analysis.
        * [SNPRelate](http://master.bioconductor.org/packages/release/bioc/html/SNPRelate.html).
          * Sample level quality control.
          * Cmputationally efficient principal component calculation.
    * [LDheatmap](http://cran.r-project.org/web/packages/LDheatmap/index.html) and [postgwas](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0071775).
        * Data visualization.
    * [plyr](http://plyr.had.co.nz/).
        * Data manipulation.
    * [genABEL](http://www.genabel.org/).
        * Statistical calculation.
    * [doParallel](http://cran.r-project.org/web/packages/doParallel/index.html).
        * parallel processing.

```{r}
# Installing necessary packages (Via console, line by line)
# ===========================================================
# source("http://bioconductor.org/biocLite.R")
# biocLite("snpStats")
# biocLite("SNPRelate")
# biocLite("rtracklayer")
# biocLite("biomaRt")
# install.packages(c('plyr', 'GenABEL', 'LDheatmap','doParallel', 'ggplot2', 'coin', 'igraph', 'devtools', 'downloader'))

# library(devtools)
# install_url("https://cran.r-project.org/src/contrib/Archive/postgwas/postgwas_1.11.tar.gz")
```

## Files and variables of interest
* Files formatted for [PLINK](http://zzz.bwh.harvard.edu/plink/).
    * .bed: binary genotype information
    * .bim: SNP's data.
        * Columns: chromosome, SNP name, genetic distance, chromosomal position, identity of allele 1 (pertains to the minor, or less common allele), identity of allele 2.
    * .fam: samples' information.
        * Columns: family ID, individual ID, paternal ID, maternal ID, sex (1 = male, 2 = female), and phenotype.
    * .CSV: A supplemental clinical file for outcome variables and additional covariates.
        * Columns: sample ID (Family ID in the .fam file), coronary artery disease status (0 = control, 1 = affected), sex (1 = male, 2 = female), age (years), triglyceride level (mg/dL), high-density lipoprotein level (mg/dL), low-density lipoprotein level (mg/dL).
```{r}
# Downloading support files
# ===========================================================

# TODO: Change to download only if it was not previously downloaded.
# Download and unzip data needed for this tutorial
# library(downloader)

# download(urlSupport, zipSupport.fn)
# unzip(zipSupport.fn, exdir = data.dir)

library(snpStats)

# Read in PLINK files
geno <- read.plink(gwas.fn$bed, gwas.fn$bim, gwas.fn$fam, na.strings = ("-9"))
```

## Data pre-processing
* The `geno` object contains a `genotype` member of type `SnpMatrix` where each column is a SNP and each row is a sample. For convenience, we assign that to the object, `genotype`.
* Filter the `genotype` data to only include samples with corresponding clinical data.

```{r}
# Obtain the SnpMatrix object (genotypes) table from geno list
# Note: Phenotypes and covariates will be read from the clinical data file, below
genotype <- geno$genotype
print(genotype) # 861473 SNPs read in for 1401 subjects

#Obtain the SNP information from geno list
genoBim <- geno$map
colnames(genoBim) <- c("chr", "SNP", "gen.dist", "position", "A1", "A2")
print(head(genoBim))

# Remove raw file to free up memory
rm(geno)

# Read in clinical file
clinical <- read.csv(clinical.fn,
                     colClasses=c("character", "factor", "factor", rep("numeric", 4)))
rownames(clinical) <- clinical$FamID
print(head(clinical))

# Subset genotype for subject data
genotype <- genotype[clinical$FamID, ]
print(genotype)  # Tutorial: All 1401 subjects contain both clinical and genotype data

# Write genotype, genoBim, clinical for future use
save(genotype, genoBim, clinical, file = working.data.fname(1))
```
### SNP level filtering
* Remove SNPs that fail to meet minimum criteria due to missing data, low variability or genotyping errors.
```{r}
# Create SNP summary statistics (MAF, call rate, etc.)
snpsum.col <- col.summary(genotype)
print(head(snpsum.col))
```
* Keep the subset of SNPs that meet minimum call rate and [MAF](https://en.wikipedia.org/wiki/Minor_allele_frequency) criterias.
```{r}
# Setting thresholds (alphas)
call <- 0.95
minor <- 0.01

# Filter on MAF and call rate
use <- with(snpsum.col, (!is.na(MAF) & MAF > minor) & Call.rate >= call)
use[is.na(use)] <- FALSE                # Remove NA's as well

cat(ncol(genotype)-sum(use),"SNPs will be removed due to low MAF or call rate.\n") #203287 SNPs will be removed

# Subset genotype and SNP summary data for SNPs that pass call rate and MAF criteria
genotype <- genotype[,use]
snpsum.col <- snpsum.col[use,]

print(genotype) # 658186 SNPs remain

# Write subsetted genotype data and derived results for future use
save(genotype, snpsum.col, genoBim, clinical, file=working.data.fname(2))

# Sample level filtering
# TODO: Fix issue: No such file or directory
# source("globals.R")

# load data created in previous snippets
load(working.data.fname(2))

library(snpStats)
```

### Basic sample filtering
* `row.summary` for sample level quality control for missing data and heterozygosity
    * Additional heterozygosity F statistic:
        * `|F|=(1−O/E)`, where:
            * O = observed proportion of heterozygous genotypes for a given sample.
            * E = expected proportion of heterozygous genotypes for a given sample, based on the MAF across all non-missing SNPs for a given sample.
```{r}
library(SNPRelate) # LD pruning, relatedness, PCA
library(plyr)

# Create sample statistics (Call rate, Heterozygosity)
snpsum.row <- row.summary(genotype)

# Add the F stat (inbreeding coefficient) to snpsum.row
MAF <- snpsum.col$MAF
callmatrix <- !is.na(genotype)
hetExp <- callmatrix %*% (2*MAF*(1-MAF))
hetObs <- with(snpsum.row, Heterozygosity*(ncol(genotype))*Call.rate)
snpsum.row$hetF <- 1-(hetObs/hetExp)

head(snpsum.row)
```
* Apply filtering on call rate and heterozygosity, selecting only those samples that meet the criteria.
```{r}
# Setting thresholds
sampcall <- 0.95 # Sample call rate cut-off
hetcutoff <- 0.1 # Inbreeding coefficient cut-off

sampleuse <- with(snpsum.row, !is.na(Call.rate) & Call.rate > sampcall & abs(hetF) <= hetcutoff)
sampleuse[is.na(sampleuse)] <- FALSE    # remove NA's as well
cat(nrow(genotype)-sum(sampleuse), "subjects will be removed due to low sample call rate or inbreeding coefficient.\n") #0 subjects removed

# Subset genotype and clinical data for subjects who pass call rate and heterozygosity crtieria
genotype <- genotype[sampleuse,]
clinical<- clinical[ rownames(genotype), ]
```
### [Identity-by-descent](https://en.wikipedia.org/wiki/Identity_by_descent) analysis
* Filter on relatedness criteria (demands GDS file format)
* `SNPRelate` package to perform IBD analysis on a subset of SNPs that are in linkage equilibrium by iteratively removing adjacent SNPs that exceed an [LD](https://en.wikipedia.org/wiki/Linkage_disequilibrium) threshold in a [sliding window](https://stackoverflow.com/a/8269948/7224879) using the `snpgdsLDpruning` function.
```{r}
# Checking for Relatedness

ld.thresh <- 0.2 # LD cut-off
kin.thresh <- 0.1 # Kinship cut-off

# Create gds file, required for SNPRelate functions
snpgdsBED2GDS(gwas.fn$bed, gwas.fn$fam, gwas.fn$bim, gwas.fn$gds)

genofile <- openfn.gds(gwas.fn$gds, readonly = FALSE)

# Automatically added "-1" sample suffixes are removed
gds.ids <- read.gdsn(index.gdsn(genofile, "sample.id"))
gds.ids <- sub("-1", "", gds.ids)
add.gdsn(genofile, "sample.id", gds.ids, replace = TRUE)

#Prune SNPs for IBD analysis
set.seed(1000)
geno.sample.ids <- rownames(genotype)
snpSUB <- snpgdsLDpruning(
  genofile,
  ld.threshold = ld.thresh,
  sample.id = geno.sample.ids, # Only analyze the filtered samples
  snp.id = colnames(genotype)) # Only analyze the filtered SNPs

snpset.ibd <- unlist(snpSUB, use.names=FALSE)
cat(length(snpset.ibd),"will be used in IBD analysis\n")  # Tutorial: expect 72812 SNPs
```
* `snpgdsIBDMoM` function computes the IBD coefficients using method of moments. The result is a table indicating kinship among pairs of samples.
```{r}
# Find IBD coefficients using Method of Moments procedure.  Include pairwise kinship.
ibd <- snpgdsIBDMoM(genofile, kinship=TRUE,
                    sample.id = geno.sample.ids,
                    snp.id = snpset.ibd,
                    num.thread = 1)

ibdcoeff <- snpgdsIBDSelection(ibd)     # Pairwise sample comparison
head(ibdcoeff)
```
* Using the IBD pairwise sample relatedness measure, iteratively remove samples that are too similar using a greedy strategy in which the sample with the largest number of related samples is removed. The process is repeated until there are no more pairs of samples with kinship coefficients above the cut-off.
```{r}
# Check if there are any candidates for relatedness
ibdcoeff <- ibdcoeff[ ibdcoeff$kinship >= kin.thresh, ]

# iteratively remove samples with high kinship starting with the sample with the most pairings
related.samples <- NULL
while ( nrow(ibdcoeff) > 0 ) {

    # count the number of occurrences of each and take the top one
    sample.counts <- arrange(count(c(ibdcoeff$ID1, ibdcoeff$ID2)), -freq)
    rm.sample <- sample.counts[1, 'x']
    cat("Removing sample", as.character(rm.sample), 'too closely related to', sample.counts[1, 'freq'],'other samples.\n')

    # remove from ibdcoeff and add to list
    ibdcoeff <- ibdcoeff[ibdcoeff$ID1 != rm.sample & ibdcoeff$ID2 != rm.sample,]
    related.samples <- c(as.character(rm.sample), related.samples)
}

# filter genotype and clinical to include only unrelated samples
genotype <- genotype[ !(rownames(genotype) %in% related.samples), ]
clinical <- clinical[ !(clinical$FamID %in% related.samples), ]

geno.sample.ids <- rownames(genotype)

cat(length(related.samples), "similar samples removed due to correlation coefficient >=", kin.thresh,"\n") 

print(genotype) # Tutorial: expect all 1401 subjects remain
```

### Ancestry
* Plot the first two principal components of the genotype data. (Achieved from `snpgdsPCA` function from `SNPRelate`).
    * Assuming homogeneity of the sample (other datasets might have to test it first), coming from european ancestry. (This is why 0 SNP will be excluded as result)
```{r}
# Checking for ancestry

# Find PCA matrix
pca <- snpgdsPCA(genofile, sample.id = geno.sample.ids,  snp.id = snpset.ibd, num.thread=1)

# TODO: Scree plot before choosing first 2 PC's
# TODO: Plot 3D using plotly
# Create data frame of first two principal comonents
pctab <- data.frame(sample.id = pca$sample.id,
                    PC1 = pca$eigenvect[,1],    # the first eigenvector
                    PC2 = pca$eigenvect[,2],    # the second eigenvector
                    stringsAsFactors = FALSE)

# Plot the first two principal comonents
plot(pctab$PC2, pctab$PC1,
     xlab="Principal Component 2",
     ylab="Principal Component 1", main = "Ancestry Plot")

# Close GDS file
closefn.gds(genofile)

# Overwrite old genotype with new filtered version
save(genotype, genoBim, clinical, file=working.data.fname(3))
```

### SNP Filtering - [HWE](https://en.wikipedia.org/wiki/Hardy%E2%80%93Weinberg_principle) filtering on control samples
* Rejection of Hardy-Weinberg equilibrium can be an indication of population substructure or genotyping errors.
    * Remove SNPs with p-values, corresponding to the HWE test statistic on CAD controls, of less than 1×10<sup>−6</sup>.
        * HWE on CAD controls due to possible violation of HWE caused by disease association.
```{r}
# Hardy-Weinberg SNP filtering on CAD controls

hardy <- 10^-6 # HWE cut-off

CADcontrols <- clinical[ clinical$CAD==0, 'FamID']
snpsum.colCont <- col.summary( genotype[CADcontrols,])
HWEuse <- with(snpsum.colCont, !is.na(z.HWE) & ( abs(z.HWE) < abs( qnorm(hardy/2))))
rm(snpsum.colCont)

HWEuse[is.na(HWEuse)] <- FALSE # Remove NA's as well
cat(ncol(genotype)-sum(HWEuse),
    "SNPs will be removed due to high HWE.\n") # 1296 SNPs removed

# Subset genotype and SNP summary data for SNPs that pass HWE criteria
genotype <- genotype[,HWEuse]
print(genotype) # 656890 SNPs remain

# Overwrite old genotype with new filtered version
save(genotype, genoBim, clinical, file=working.data.fname(4))
```